---
title: "00_simulated_data"
author: "Frances Lin"
date: "11/11/2020"
output: pdf_document
---

### Description

This .Rmd file creates simulated data for the two-component Gaussian mixture. All related results can be found in the results folder. 

### Load packages

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(tibble)
library(ggplot2)
library(gridExtra)
```

### Simulate data

Specify the values and parameters for the two-component Gaussian mixture
```{r}
# Specify the #s of obversations for each component
n <- 100
k <- 2 # might want to extend it to kth mixture

# Specify mean and sd of component 1 (or A)
mu_1 <- 0
sd_1 <- 1

# Specify mean and sd of component 2 (or B)
mu_2 <- 2 #4 #2
sd_2 <- 1
```

Create a dataframe to store the simulated data
```{r}
set.seed(1234) # for reproducibility
data <- tibble(
  component = c(rep("A", n), rep("B", n)), 
  value = c(rnorm(n, mean = mu_1, sd = sd_1), rnorm(n, mean = mu_2, sd = sd_2)) 
)
head(data)
```

### Plot simulated data

```{r}
data_flat_plot <- data %>%
  ggplot(aes(x = value, y = "A", color = factor(component))) +
  geom_point(alpha = 0.4) + 
  ggtitle("Simulated Data -- Complete Dataset")
#data_flat_plot
```

```{r}
data_density_plot <- data %>%
  ggplot(aes(x = value, color = factor(component))) +
  geom_density()
#data_density_plot
```

```{r}
data_plot <- grid.arrange(data_flat_plot, data_density_plot)
data_plot
```

Estimate the mean, var, and sd for this complete dataset (component A and B) using MLE
```{r}
data_summary <- data %>%
  group_by(component) %>%
  summarize(
    mean = mean(value), 
#    var = var(value), # never mind dnorm() takes sd
    var = var(value),
    sd = sd(value)
  )
data_summary
```

### Plot simulated data

Assume there is a latent (or missing) variable
```{r}
data_flat_plot_latent <- data %>%
  ggplot(aes(x = value, y = "A")) +
  geom_point(alpha = 0.4) + 
  ggtitle("Simulated Data -- Incomplete Dataset")
#data_flat_plot_latent
```

```{r}
data_density_plot_latent <- data %>%
  ggplot(aes(x = value)) +
  geom_density()
#data_density_plot_latent
```

```{r}
data_plot_latent <- grid.arrange(data_flat_plot_latent, data_density_plot_latent)
data_plot_latent
```

We'll see how do we estimate the mean and sd for this incomplete dataset (with components removed) next 

### Save out results

```{r}
write_rds(data, here("results", "data.rds"))
write_rds(data_plot, here("results", "data_plot.rds"))
write_rds(data_summary, here("results", "data_summary.rds"))
write_rds(data_plot_latent, here("results", "data_plot_latent.rds"))
```

