---
title: "01_initialization_kmeans"
author: "Frances Lin"
date: "11/11/2020"
output: pdf_document
---

### Description

This .Rmd file calculates the initial estimates for the first EM iteration using K-means clustering. All related results can be found in the results folder. 

### Load packages

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(tibble)
library(stats)
```

The author of this [article](https://tinyheero.github.io/2016/01/03/gmm-em.html#references) suggests that it is common to use K-means clustering (hard labelling) to obtain the initial estimates for EM algorithm (the soft labelling).

### Load data

```{r}
data <- readRDS(here("results", "data.rds"))
head(data)
```

### Perform k-mean clustering

```{r}
# K-means clustering is "hard" because we assign k = 2 
# (i.e. a data point is either 1 or 2)
value <- data$value
kmeans <- kmeans(x = value, centers = 2) 
kmeans_cluster <- kmeans$cluster
```

### Plot k-mean clustering

```{r}
df_kmeans <- tibble(
  value = value, 
  cluster = kmeans_cluster
)
head(df_kmeans)
```

```{r}
df_plot_kmeans <- df_kmeans %>%
  mutate(data_number = row_number()) %>%
  ggplot(aes(x = value, y = data_number, color = factor(cluster))) +
  geom_point() + 
  ggtitle("K-means Clustering")
df_plot_kmeans
```

### Store the values from K-means clustering, which then become initial estimates for EM algorithm

Estimate mean $(\mu_1, \mu_2)$ and sd $(\sigma_1, \sigma_2)$ 

```{r}
df_summary_kmeans <- df_kmeans %>%
  group_by(cluster) %>%
  summarize(
    mean = mean(value), 
    var = var(value), 
    sd = sd(value),
    size = length(value)
  )
df_summary_kmeans
```

Estimate weighting probability $\pi$ 

```{r}
df_summary_kmeans <- df_summary_kmeans %>% 
  mutate(
    pi = size / sum(size))
df_summary_kmeans
```

Now we can pass the values to EM algorithm

### Save out results

```{r}
write_rds(df_kmeans, here("results", "df_kmeans.rds"))
write_rds(df_plot_kmeans, here("results", "df_plot_kmeans.rds"))
write_rds(df_summary_kmeans, here("results", "df_summary_kmeans.rds"))
```

